Modeling the Latent Space of Symbolic String Quartet Music

### 2021-06-06 papers/final/alex/paper.tex
hparam tuning for vae model

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}

@@

@@

MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}. This model is trained on 16-bar
paragraphs of music and is capable of generating new melodies that
blend two given melodies via latent space interpolation.

Another Magenta model called Music Transformer is a generative model
that borrows its approach from the Natural Language Processing (NLP)
domain, using a self-attention network to model MIDI music as a
sequence of discrete tokens with relative positional dependencies
\cite{huang_music_2018}. The focus of this model is on learning
long-term dependencies in music to produce longer clips of music with
coherent structure. Music Transformer was trained on a dataset of
Piano-e-competition performances \cite{hawthorne2019enabling} and its
generated piano music received favorable qualitative (Likert scale)
ratings from human listeners for its resemblance to human-composed
music \cite{huang_music_2018}.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{song_structure.png}
  \caption{Diagram of the hierarchical structure of a musical
    composition, from the MuseGAN paper \cite{dong2017musegan}.}
  \label{song_structure}
  \Description{A diagram showing the hierarchical structure of songs,
    to paragraphs, to phrases, to bars, to, beats, to pixels.}
\end{figure}

@@

@@

In order to accommodate polyphonic music, we each MIDI file is
converted into a modified pianoroll representation, an example of
which is visualized in Figure \ref{pianoroll}. In this modified
representation, each instrument track is a sequence of integers from
0-129 representing the MIDI note pitch value being played at the given
time step, where values 0-127 are pitches, 128 is a rest (silence) and
129 is sustain, indicating that the previously played pitch is held
continuously. Because chords (multiple notes being played
simultaneously by the same instrument) are relatively rare in string
quartets, the lowest note of each chord is taken and the rest
discarded, to reduce dimensionality. Because notes at the extreme low
and high end are rare and are out of range of common instruments, the
note pitch values are clipped to a narrower range and ordinal
encoded. For the 36 string quartets found in the MusicNet database,
this results in 66 distinct values, where 0-63 represent notes played
by the violins, viola and cello, 64 is rest and 65 is sustain. Our
software package also includes functions to convert this
representation back into a Music21 Score object, which can then be
written to a MIDI file.

Because songs are typically at least a few minutes long and of varying
length, it will not be feasible to train with entire songs as
examples, so we will crop songs into phrases of equal numbers of
measures to use as training data.

@@

The model is a recurrent variational autoencoder; both Long Short-Term
Memory (LSTM) and Gated Recurrent Unit (GRU) cell types were tested.

@@

The decoder model samples from the latent space $z$ with the
distribution parameterized by the $\mu$ and $log(\sigma)$ values
learned by the encoder.

@@

>>>>

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,authorversion]{acmart}
%% NOTE that a single column version may be required for
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to
%% \documentclass[manuscript,screen,review]{acmart}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

A survey of related works in generative modeling of polyphonic music
turned up several models focused on multi-instrument pop/rock music
arrangements or classical piano performances. In contrast, this
project focuses on variational autoencoding and latent space
interpolation of classical string quartets (arrangements for an
ensemble of two violins, one viola and one cello) and is, to the
author's knowledge, the first project to apply this modeling approach
to this type of musical arrangement.

@@

MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent codes and rely on autoregression to
generate output sequences \cite{roberts_hierarchical_2018}. This model
is trained on 16-bar paragraphs of music and is capable of generating
new melodies that blend two given melodies together via latent space
interpolation. Among the literature surveyed, MusicVAE's approach is
the most similar to this work; in comparison our model is
significantly simpler and works on a single bar of music at a time.

Another Magenta model called Music Transformer is a generative model
that borrows its approach from the Natural Language Processing (NLP)
domain, using a self-attention network to model MIDI music as a
sequence of discrete tokens with relative positional dependencies
\cite{huang_music_2018}. The focus of this model is on learning
long-term dependencies in music to produce longer clips of music with
coherent structure. Music Transformer was trained on a dataset of
Piano-e-competition performances \cite{hawthorne2019enabling} and its
generated piano music received favorable qualitative (Likert scale)
ratings from human listeners for its resemblance to human-composed
music \cite{huang_music_2018}. In contrast to MusicVAE and this work,
Music Transformer is capable of generating much longer sequences of
music, but does so via sequence extrapolation rather than
interpolation, so it can continue a given priming melody but does not
blend multiple given melodies together. It also assumes a single
instrument track (piano).

@@

@@

While there may be significant opportunity in future approaches that
combining the two, contemporary research generally treats symbolic and
audio music modeling as separate problem domains and this work does
the same, choosing to focus on symbolic music.

@@

In order to accommodate polyphonic music, we each MIDI file is
converted into a modified pianoroll representation, an example of
which is visualized in Figure \ref{pianoroll}. In this modified
representation, each instrument track is a sequence of integers from
0-129 representing the MIDI note pitch value being played at the given
time step, where values 0-127 are pitches, 128 is a rest (silence) and
129 is sustain, indicating that the previously played pitch is held
continuously. MIDI data contains pressure values that indicate how
hard or soft each note is played; this information is discarded in our
preprocessing and all notes are represented with constant
pressure. Because chords (multiple notes being played simultaneously
by the same instrument) are relatively rare in string quartets, the
lowest note of each chord is taken and the rest discarded, to reduce
dimensionality. Because notes at the extreme low and high end are rare
and are out of range of common instruments, the note pitch values are
clipped to a narrower range and ordinal encoded. For the 36 string
quartets found in the MusicNet database, this results in 66 distinct
values, where 0-63 represent notes played by the violins, viola and
cello, 64 is rest and 65 is sustain. Our software package also
includes functions to convert this representation back into a Music21
Score object, which can then be written to a MIDI file.

Because songs are typically at least a few minutes long and of varying
length, it will not be feasible to train with entire songs as
examples, so we will crop songs into phrases of equal numbers of
beats to use as training data.

@@

The model is a recurrent variational autoencoder; both Long Short-Term
Memory (LSTM) and Gated Recurrent Unit (GRU) cell types were tested
but this choice did not make a significant difference in the results.

@@

The decoder model samples from the latent space $z$ with the
distribution parameterized by the $\mu$ and $log(\sigma)$ values
learned by the encoder, and then splits into four parallel output
paths, one per instrument track. Each path contains again two LSTM
layers with a dropout layer between them, and then a fully connected
layer with softmax activation is used to select the most probable note
value for the current track at the current time step.

The autoencoder reconstruction loss function is sparse categorical
cross-entropy, which is added to the Kullback-Leibler divergence of
the approximate posterior distribution from a multivariate Gaussian
distribution with mean 0 and standard deviation 1, resulting in the
(negative) evidence lower bound (ELBO) loss, which is a standard
variational autoencoder loss function. This loss function is applied
to each track independently and then summed across all four tracks
with equal weight.

@@

\subsection{Model Experimentation}

The following hyperparameters were tuned, with tested values listed
and best parameters in \textbf{bold}:

\begin{itemize}
  \tightlist
  \item batch size: \textbf{32}, 64, 96
  \item learning rate: 0.001, \textbf{0.0002}
  \item recurrent layer type: LSTM, GRU (made no difference)
  \item recurrent encoder layer directions: \textbf{unidirectional}, bidirectional
  \item recurrent layer cells: 64, \textbf{128}, 256
  \item latent code dimension: 128, 256, \textbf{512}
  \item embedding dimension: 4, \textbf{8}, 16
  \item dropout rate: 0.2, 0.4, \textbf{0.5}
\end{itemize}

Each model was trained for up to 500 epochs with an early stopping
patience of 20 epochs. The model was cross-validated on 10\% of the
training data after each epoch. The best model was selected by a
combination of validation loss and listening to hand-picked sample
reconstructions and interpolations.

The reconstructed music produced by the model sounds reminiscent of
the original inputs, preserving the key signature and general rhythmic
and melodic structure, albeit with noticeable differences.

A website with several samples of the original input MIDI tracks, the
autoencoded outputs, and interpolations, is presented at: TODO

TODO: links to audio samples
